---
title: "Health Analytics Collaboratory (HAC) - University of Michigan"
subtitle: "<h2><u>HAC Alzheimer's Disease Digital Health Analytics Project: Loading and Converting Data Files </u></h2>"
author: "<h3>Ivo Dinov, Cooper Stansbury, Haiyin Liu, Bingxin Chen</h3>"
date: "`r format(Sys.time(), '%B %Y')`"
tags: [HAC, DSPA, SOCR, MIDAS, Big Data, Predictive Analytics, Computable Phenotypes] 
output: 
  html_notebook
  df_print: paged
---

# Introduction

This notebook serves as an introduction to the data. This notebook covers data loading (and saving) and basic descriptive peeks. 

## Session Information and Configuration

```{r}
library(DataExplorer)
library(lemon)
```


This section outputs information regarding loaded dependencies and OS configuration that may help in trouble-shooting. 

```{r}
sessionInfo()
```

```{r}
sapply(c('repr', 'IRdisplay', 'IRkernel'), function(p) paste(packageVersion(p)))
```

Check the working directory. Expected to be a the 'project' level for knit. 

```{r}
getwd()
```


# Data Loading
This section contains code "blocks"" for saving data in different file formats for ingestion into different downstream programs. The source code for these functions is in the file `code/load_and_convert_functions.R` for re-use in analysis notebooks. The following function loads data from the `code/load_and_convert_functions.R` file to the current workspace.

```{r}
source('code/load_and_convert_functions.R')
```

## Sample Data

An example of loading the sample data into a `data.frame` from a MS Excel file:

```{r}
# df <- read_xlsx_from_path(filepath = '../../local_data/DS_PreprocessedAnonymized_Text.xlsx')
```


## Demographics

Local path specification:

```{r include=FALSE}
demographics_path <- '../../local_data/full_data/Demographics.dsv'
```


```{r}
demo_df <- read_multi_character_delim_fast(demographics_path, v=FALSE, "<break>", header = TRUE)
nrow(demo_df) #Expect 2052 total records (first record is a header).
```

```{r}
head(demo_df)
```

```{r}
knit.print.data.frame <- lemon_print
```


```{r}
str(demo_df)
```


```{r}
summary(demo_df)
```

```{r}
introduce(demo_df)
```

```{r}
plot_bar(demo_df)
```



```{r}
plot_bar(demo_df$AGE_CUR_OR_AT_DEATH)
```

```{r}
plot_bar(demo_df$EDUCATION_YEARS)
```

```{r}
plot_histogram(as.numeric(demo_df$AGE_CUR_OR_AT_DEATH))
```


## Encounters
Expect 304573 records (first record is a header).

```{r}
encounters_path <- '../../local_data/full_data/Encounters.dsv'
```


```{r}
gc()
enc_df <- read_multi_character_delim_fast(encounters_path, "<break>", header = TRUE)
nrow(enc_df)
```






## Labs
It looks like 'Labs' is already a `.csv` file. Expect 1516205 records (first record is a header).

```{r}
labs_path <- '../../local_data/full_data/Labs.dsv'
```


```{r}
labs_df <- read.csv2(labs_path)
```







## Notes
Expect 11078079 records (first record is a header).

**WARNING:** the following file is magnitudes larger than the others and requires a long time (and a lot of memory) to load. Proceed with caution.

```{r}
notes_path <- '../../local_data/full_data/Notes.dsv'
```



```{r}
gc()
source('code/load_and_convert_functions.R')
note_df <- read_notes_custom_delim(notes_path, v=FALSE, '<break>', header = TRUE)
nrow(note_df)
```









## Radiology 
Expect 2792 records (first record is a header).

```{r}
radiology_path <- '../../local_data/full_data/Radiology.dsv'
```


```{r}
gc()
radio_df <- read_multi_character_delim_fast(radiology_path, v=FALSE, "<break>", header = TRUE)
nrow(radio_df)
```














#Apppendix: Custom Load and Convert Functions 

Below is a bash shell commad to print the functions in `code/load_and_convert_functions.R` to the console:

```{bash}
# cat code/load_and_convert_functions.R
```

# Appendix: Saving Data

## R Objects
R changes rapidly. File types specific to R may not be robust to future changes in technology. ** Note on `save`:** the R function `save` stores both the object and the name of the object. This is important when loading objects back into R. `.Rdata` (sometimes `.Rda`) files are specific to the R environment. They can store arbitrary numbers of objects in a single file. For more documentation I suggest: 
[YaRrr/rdata-files](https://bookdown.org/ndphillips/YaRrr/rdata-files.html).

**Note for OSX Users:** The extension may matter. We have chosen `.Rdata`. Reference: [error-in-gzfilefile-wb-cannot-open-the-connection-or-compressed-file](https://stackoverflow.com/questions/15363076/error-in-gzfilefile-wb-cannot-open-the-connection-or-compressed-file).

The following block demonstrates saving both `.Rdata` or `.rds` format objects.

```{r}
# save_rData(df, 'sample.Rdata')
# save_rData(df, 'sample.rds')
```


## CSV
This is the most robust long-term file storage format. Note that there are several possible arguments for `write.csv()` that may be necessary to specify depending on downstream requirements.

```{r}
#  save_csv(df, '../../local_data/', 'mini_sample_test.csv')
```

```{r}
# save_csv(demo_df, '../../local_data/stuctured_csv_data/', 'demographics_test.csv')
# save_csv(enc_df, '../../local_data/stuctured_csv_data/', 'enounters_test.csv')
# save_csv(lab_df, '../../local_data/stuctured_csv_data/', 'labs_test.csv')
# save_csv(radio_df, '../../local_data/stuctured_csv_data/', 'radiology_test.csv')
```


Please recall that this data object is much larger than the others. 

```{r}
# save_csv(note_df, '../../local_data/structured_csv_data/', 'notes_test.csv')
```




